---
title: "STA2016 Final Project"
author: "Nevena Ciganovic"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

# Libraries used throughout
library(dplyr); library(readr); library(tidyr); library(stringr); library(janitor)
library(sf); library(leaflet); library(htmltools); library(scales)

# Key cleaner used for joining names
clean_key <- function(x) {
  x |>
    as.character() |> str_squish() |>
    stringi::stri_trans_general("Latin-ASCII") |>
    tolower() |>
    str_replace_all("&", " and ") |>
    str_replace_all("[^a-z0-9]+", "_") |>
    str_replace_all("^_|_$", "")
}

```

## Proposal Code

```{r build_points_and_polys, warning=F, message=F, include=F}
setwd("/Users/nevenaciganovic/Downloads/STA2016/project")

# --- Shootings (points) --------------------------------------------
shoot <- read_csv(
  "Shooting_and_Firearm_Discharges_Open_Data_-2630709695104719865.csv",
  show_col_types = FALSE
)

stopifnot(all(c("LONG_WGS84","LAT_WGS84") %in% names(shoot)))

shoot <- shoot |>
  mutate(
    LONG_WGS84 = suppressWarnings(as.numeric(LONG_WGS84)),
    LAT_WGS84  = suppressWarnings(as.numeric(LAT_WGS84))
  ) |>
  filter(
    is.finite(LONG_WGS84), is.finite(LAT_WGS84),
    LONG_WGS84 > -90, LONG_WGS84 < -70,
    LAT_WGS84  >  40, LAT_WGS84  <  50
  )

shoot_sf <- st_as_sf(
  shoot, coords = c("LONG_WGS84","LAT_WGS84"),
  crs = 4326, remove = FALSE
)

# --- Toronto Neighbourhoods (polygons) -----------------------------
cand_paths <- c(
  "Toronto_Neighbourhoods/Toronto_Neighbourhoods.shp",
  "Toronto_Neighbourhoods.shp",
  "~/Downloads/Toronto_Neighbourhoods/Toronto_Neighbourhoods.shp"
)
shp_path <- cand_paths[file.exists(path.expand(cand_paths))][1]
if (is.na(shp_path)) stop("Could not find the Toronto Neighbourhoods shapefile.")

neighbourhood <- st_read(shp_path, quiet = TRUE) |>
  st_transform(4326)

# Your shapefile's name field:
name_col <- "Neighbourh"
stopifnot(name_col %in% names(neighbourhood))

# Add clean join key + pretty display name
neighbourhood <- neighbourhood |>
  mutate(
    demo_name   = clean_key(.data[[name_col]]),
    pretty_name = str_to_title(str_replace_all(demo_name, "_", " "))
  )
```

```{r build_demo_keep,warning=F, message=F, include=FALSE}
setwd("/Users/nevenaciganovic/Downloads/STA2016/project")

# --- 2021 Neighbourhood Profiles → demo_keep -----------------------
profiles <- read_csv("neighbourhood-profiles-2021-158-model.csv", show_col_types = FALSE) |>
  clean_names()

neigh_cols <- setdiff(names(profiles), "neighbourhood_name")

long <- profiles |>
  pivot_longer(all_of(neigh_cols), names_to = "neigh", values_to = "value_raw") |>
  mutate(
    indicator = str_to_lower(str_squish(neighbourhood_name)),
    neigh     = str_to_lower(str_squish(neigh)),
    value     = readr::parse_number(as.character(value_raw))
  )

unemp_tbl <- long |>
  filter(str_detect(indicator, "^unemployment rate$")) |>
  transmute(neigh, unemploy_rate = value)

inc_tbl <- long |>
  filter(str_detect(indicator, "^median total income of household in 2020")) |>
  transmute(neigh, median_total_income = value)

vismin_total <- long |>
  filter(str_detect(indicator, "^total visible minority population$")) |>
  transmute(neigh, vismin_total = value)

not_vismin <- long |>
  filter(str_detect(indicator, "^not a visible minority$")) |>
  transmute(neigh, not_vismin = value)

black_tbl <- long |>
  filter(str_detect(indicator, "^black$")) |>
  transmute(neigh, black_count = value)

race <- vismin_total |>
  full_join(not_vismin, by = "neigh") |>
  full_join(black_tbl,  by = "neigh") |>
  mutate(
    population_total       = vismin_total + not_vismin,
    pct_visible_minority   = if_else(population_total > 0, 100 * vismin_total / population_total, NA_real_),
    pct_black              = if_else(population_total > 0, 100 * black_count   / population_total, NA_real_)
  ) |>
  select(neigh, population_total, pct_visible_minority, pct_black)

demo_keep <- race |>
  left_join(inc_tbl,  by = "neigh") |>
  left_join(unemp_tbl, by = "neigh") |>
  rename(demo_name = neigh) |>
  mutate(demo_name = clean_key(demo_name)) |>
  distinct(demo_name, .keep_all = TRUE)
```

```{r warning=F, message=F, echo=F}
# --- Setup ---------------------------------------------------------
setwd("/Users/nevenaciganovic/Downloads/STA2016/project")

library(dplyr)
library(readr)
library(sf)
library(leaflet)
library(htmltools)

# --- Load shootings (no year filtering) ---------------------------
shoot <- read_csv(
  "Shooting_and_Firearm_Discharges_Open_Data_-2630709695104719865.csv",
  show_col_types = FALSE
)

stopifnot(all(c("LONG_WGS84", "LAT_WGS84") %in% names(shoot)))
shoot <- shoot |>
  filter(
    LONG_WGS84 > -90 & LONG_WGS84 < -70,
    LAT_WGS84  >  40 & LAT_WGS84  <  50
  )

shoot_sf <- st_as_sf(
  shoot,
  coords = c("LONG_WGS84", "LAT_WGS84"),
  crs = 4326,
  remove = FALSE
)

# --- Load Toronto Neighbourhoods (polygons) -----------------------
cand_paths <- c(
  "Toronto_Neighbourhoods/Toronto_Neighbourhoods.shp",
  "Toronto_Neighbourhoods.shp",
  "~/Downloads/Toronto_Neighbourhoods/Toronto_Neighbourhoods.shp"
)
shp_path <- cand_paths[file.exists(path.expand(cand_paths))][1]
if (is.na(shp_path)) stop("Could not find the Toronto Neighbourhoods shapefile.")

neighbourhood <- st_read(shp_path, quiet = TRUE) |> st_transform(4326)

# Prefer a NAME-like field (not IDs)
name_candidates <- c(
  "Neighbourh","NEIGHBOURH","NEIGHBOURHOOD","NEIGHBOURHOOD_NAME",
  "AREA_NAME","AREA_NAME_1","NAME","Name"
)
nm_col <- intersect(name_candidates, names(neighbourhood))
if (length(nm_col) == 0) {
  # fallback to first non-geometry column
  nm_col <- setdiff(names(neighbourhood), attr(neighbourhood, "sf_column"))[1]
} else {
  nm_col <- nm_col[1]
}

lab_neigh <- ~htmlEscape(get(nm_col))

# --- Map with layer toggles ---------------------------------------
leaflet(options = leafletOptions(preferCanvas = TRUE)) |>
  addTiles(group = "Base") |>

  # Neighbourhood boundaries (hover shows NAME)
  addPolygons(
    data = neighbourhood, group = "Neighbourhoods",
    label = lab_neigh,
    weight = 1, color = "#666666",
    fillColor = "#e6f2ff", fillOpacity = 0.15,
    highlightOptions = highlightOptions(weight = 2, color = "#333333", bringToFront = TRUE),
    labelOptions = labelOptions(
      style = list("font-weight" = "600"),
      textsize = "13px",
      direction = "auto"
    )
  ) |>

  # Shooting points
  addCircleMarkers(
    data  = shoot_sf, group = "Shootings",
    radius = 4, stroke = FALSE, fillOpacity = 0.7,
    fillColor = "red"
  ) |>

  # Layer control
  addLayersControl(
    baseGroups    = c("Base"),
    overlayGroups = c("Neighbourhoods", "Shootings"),
    options = layersControlOptions(collapsed = FALSE)
  ) |>

  # Fit to neighbourhood extent (stable and tidy)
  fitBounds(
    lng1 = st_bbox(neighbourhood)[["xmin"]], lat1 = st_bbox(neighbourhood)[["ymin"]],
    lng2 = st_bbox(neighbourhood)[["xmax"]], lat2 = st_bbox(neighbourhood)[["ymax"]]
  )

```

```{r message=FALSE, warning=FALSE, echo=F}
library(dplyr)
library(sf)
library(stringr)
library(stringi)
library(tidyr)
library(leaflet)
library(htmltools)
library(scales)

# --- 1) One cleaning function used everywhere ----------------------
clean_key <- function(x) {
  x %>%
    as.character() %>%
    str_squish() %>%
    stringi::stri_trans_general("Latin-ASCII") %>%  # strip accents
    tolower() %>%
    str_replace_all("&", " and ") %>%
    str_replace_all("[^a-z0-9]+", "_") %>%
    str_replace_all("^_|_$", "")
}

# --- 2) Build keys on BOTH sides ----------------------------------
name_col <- "Neighbourh"   # from your shapefile
neighbourhood <- neighbourhood %>%
  mutate(
    demo_name   = clean_key(.data[[name_col]]),
    pretty_name = str_to_title(str_replace_all(demo_name, "_", " "))
  )

demo_keep <- demo_keep %>%
  mutate(demo_name = clean_key(demo_name)) %>%   # ensure same key format
  distinct(demo_name, .keep_all = TRUE)

# --- 3) Join demographics to polygons -----------------------------
neigh_demo <- neighbourhood %>%
  left_join(demo_keep, by = "demo_name")

# Optional: fallback to shapefile population if CSV didn’t match
if ("Total_Popu" %in% names(neigh_demo)) {
  neigh_demo <- neigh_demo %>%
    mutate(population_total = coalesce(population_total, as.numeric(Total_Popu)))
}

# --- 4) Count shootings by neighbourhood (spatial join) -----------
shoot_sf   <- st_transform(shoot_sf, st_crs(neigh_demo))
neigh_demo <- st_make_valid(neigh_demo)

shoot_counts <- st_join(
  shoot_sf,
  neigh_demo["demo_name"],
  join = st_within,
  left = FALSE
) %>%
  st_drop_geometry() %>%
  count(demo_name, name = "shootings_n")

neigh_demo <- neigh_demo %>%
  left_join(shoot_counts, by = "demo_name") %>%
  mutate(
    shootings_n  = replace_na(shootings_n, 0),
    rate_per_10k = if_else(is.finite(population_total) & population_total > 0,
                           10000 * shootings_n / population_total,
                           NA_real_)
  )

# --- 5) Choropleth (quintiles) ------------------------------------
pal <- colorQuantile(
  palette  = "YlOrRd",
  domain   = neigh_demo$rate_per_10k,
  n        = 5,
  na.color = "#BDBDBD"
)

popup_poly <- sprintf(
  "<b>%s</b><br/>Shootings: %s<br/>Population: %s<br/>Rate/10,000: %s",
  neigh_demo$pretty_name,
  comma(neigh_demo$shootings_n),
  ifelse(is.na(neigh_demo$population_total), "NA", comma(neigh_demo$population_total)),
  ifelse(is.finite(neigh_demo$rate_per_10k), sprintf('%.2f', neigh_demo$rate_per_10k), "NA")
) |> lapply(HTML)

leaflet(neigh_demo, options = leafletOptions(preferCanvas = TRUE)) |>
  addTiles() |>
  addPolygons(
    fillColor   = ~pal(rate_per_10k),
    fillOpacity = 0.75,
    color       = "#555555",
    weight      = 1,
    opacity     = 0.9,
    label       = ~pretty_name,   # name on hover
    labelOptions = labelOptions(textsize = "12px", direction = "auto",
                                style = list("font-weight" = "600")),
    popup       = popup_poly,
    highlightOptions = highlightOptions(weight = 2, color = "#000", bringToFront = TRUE)
  ) |>
  addLegend(
    pal    = pal,
    values = ~rate_per_10k,
    title  = "Shootings per 10,000 (quintiles)",
    position = "bottomright",
    opacity = 0.8,
    labFormat = labelFormat(digits = 2)
  ) |>
  fitBounds(
    lng1 = st_bbox(neigh_demo)[["xmin"]], lat1 = st_bbox(neigh_demo)[["ymin"]],
    lng2 = st_bbox(neigh_demo)[["xmax"]], lat2 = st_bbox(neigh_demo)[["ymax"]]
  )


```

```{r message=F, warning=F, echo=F}
setwd("/Users/nevenaciganovic/Downloads/STA2016/project")

library(dplyr)
library(readr)
library(tidyr)
library(stringr)
library(janitor)
library(gt)
library(scales)

profiles <- read_csv("neighbourhood-profiles-2021-158-model.csv", show_col_types = FALSE) |>
  clean_names()

# Wide → long
neigh_cols <- setdiff(names(profiles), "neighbourhood_name")
long <- profiles |>
  pivot_longer(all_of(neigh_cols), names_to = "neigh", values_to = "value_raw") |>
  mutate(
    indicator = str_to_lower(str_squish(neighbourhood_name)),
    neigh     = str_to_lower(str_squish(neigh)),
    value     = readr::parse_number(as.character(value_raw))
  )

# --- Exact row picks based on your grep output ---
# Unemployment rate
unemp_tbl <- long |>
  filter(str_detect(indicator, "^unemployment rate$")) |>
  transmute(neigh, unemploy_rate = value)

# Median household income (2020) — per 2021 Census
inc_tbl <- long |>
  filter(str_detect(indicator, "^median total income of household in 2020")) |>
  transmute(neigh, median_total_income = value)

# Visible minority components (to build population + %s)
vismin_total <- long |>
  filter(str_detect(indicator, "^total visible minority population$")) |>
  transmute(neigh, vismin_total = value)

not_vismin <- long |>
  filter(str_detect(indicator, "^not a visible minority$")) |>
  transmute(neigh, not_vismin = value)

black_tbl <- long |>
  filter(str_detect(indicator, "^black$")) |>
  transmute(neigh, black_count = value)

# Build population (private households) + %VM + %Black
race <- vismin_total |>
  full_join(not_vismin, by = "neigh") |>
  full_join(black_tbl, by = "neigh") |>
  mutate(
    population_total       = vismin_total + not_vismin,
    pct_visible_minority   = if_else(population_total > 0, 100 * vismin_total / population_total, NA_real_),
    pct_black              = if_else(population_total > 0, 100 * black_count   / population_total, NA_real_)
  ) |>
  select(neigh, population_total, pct_visible_minority, pct_black)

# Final demographics table for join
demo_keep <- race |>
  left_join(inc_tbl,  by = "neigh") |>
  left_join(unemp_tbl, by = "neigh") |>
  rename(demo_name = neigh) |>
  select(demo_name, population_total, median_total_income, unemploy_rate, pct_visible_minority, pct_black)

# Make pretty table
demo_tbl <- demo_keep %>%
  # Keep only needed columns in the right order
  select(
    neighbourhood = demo_name,
    population_total,
    median_total_income,
    unemploy_rate,
    pct_visible_minority,
    pct_black
  ) %>%
  # Fix names (underscores → spaces, title case)
  mutate(
    neighbourhood = str_replace_all(neighbourhood, "_", " "),
    neighbourhood = str_to_title(neighbourhood)
  )

gt(demo_tbl) %>%
  tab_header(
    title = md("**Toronto Neighbourhood Demographics (2021)**"),
    subtitle = md("Selected indicators from the 2021 Census")
  ) %>%
  cols_label(
    neighbourhood        = "Neighbourhood",
    population_total     = "Population",
    median_total_income  = "Median HH Income (2020, $)",
    unemploy_rate        = "Unemployment Rate (%)",
    pct_visible_minority = "Visible Minority (%)",
    pct_black            = "Black (%)"
  ) %>%
  fmt_number(columns = c(population_total), sep_mark = ",", decimals = 0) %>%
  fmt_currency(columns = c(median_total_income), currency = "CAD", decimals = 0) %>%
  fmt_number(columns = c(unemploy_rate, pct_visible_minority, pct_black), decimals = 1) %>%
  tab_spanner(
    label = "Race / Ethnicity",
    columns = c(pct_visible_minority, pct_black)
  ) %>%
  opt_row_striping() %>%
  tab_options(
    table.width = px(900),
    data_row.padding = px(6)
  ) %>%
  tab_source_note(md("Source: City of Toronto Neighbourhood Profiles (2021)."))

```

## RQ1: Are firearm violence incidents spatially clustered across Toronto, or consistent with complete spatial randomness (CSR)?


```{r packages, message=FALSE, warning=FALSE, echo=FALSE}
set.seed(123)
library(sf)
library(dplyr)
library(spatstat.geom)
library(spatstat.explore)
library(spatstat.model)
```

### Build a `ppp` point pattern in meters

```{r, message=FALSE, warning=FALSE, echo=FALSE}
# Project to meters (Toronto: UTM 17N is usually fine)
crs_m <- 26917

W_sf <- neighbourhood |> st_transform(crs_m) |> st_union() |> st_as_sf()
X_sf <- shoot_sf      |> st_transform(crs_m)

# Convert to spatstat window + points
W_owin <- as.owin(W_sf)
xy     <- st_coordinates(X_sf)

# Keep only points inside window
inside <- inside.owin(xy[,1], xy[,2], W_owin)
xy_in  <- xy[inside, , drop = FALSE]

# Create ppp object
X_ppp <- ppp(x = xy_in[,1], y = xy_in[,2], window = W_owin)

# Jitter duplicated points by a tiny amount (1 meter)
X_ppp <- rjitter(X_ppp, retry = TRUE, radius = 1)

# Check duplicates are gone
any(duplicated(cbind(X_ppp$x, X_ppp$y)))

# Plot
plot(X_ppp,
     main = "Point Pattern of Shootings (projected, meters)",
     pch = 16, cex = 0.4)

```

### Are incidents clustered beyond complete spatial randomness?

Test for CSR using Ripley's K and G function.

```{r}
library(spatstat.explore)
set.seed(123)

nsim <- 39

# -----------------------------
# L(r)-r CSR envelope
# -----------------------------
rmax <- diameter(Window(X_ppp))/4
rvec <- seq(0, rmax, length.out = 64)

L_env <- envelope(
  X_ppp,
  fun = Lest,
  nsim = nsim,
  global = TRUE,
  r = rvec,
  correction = "border",
  verbose = FALSE
)

plot(L_env, . - r ~ r,
     main = "CSR test: L(r) - r (global Monte Carlo envelope)",
     xlab = "r (meters)",
     ylab = "L(r) - r",
     legend = FALSE)
abline(h = 0, lty = 2)

# -----------------------------
# G(r) CSR envelope
# (choose r based on NN distances)
# -----------------------------
nn <- nndist(X_ppp)
rmax_g <- as.numeric(quantile(nn, 0.95))     # show up to 95% of NN distances
rvec_g <- seq(0, rmax_g, length.out = 64)

G_env <- envelope(
  X_ppp,
  fun = Gest,
  nsim = nsim,
  global = TRUE,
  r = rvec_g,
  correction = "rs",
  verbose = FALSE
)

plot(G_env,
     main = "CSR test: G(r) (global Monte Carlo envelope)",
     xlab = "r (meters)",
     ylab = "G(r)",
     legend = FALSE)


```

Formal MAD test.
```{r}
library(spatstat.explore)

mad_out <- mad.test(
  X_ppp, fun = Lest,
  nsim = nsim,
  rinterval  = c(0, rmax),
  correction = "border",
  simulate   = expression(runifpoint(npoints(X_ppp), win = X_ppp$window))
)
mad_out

```

Conclusion: CSR is rejected. There is clear spatial clustering under CSR.


```{r}
library(spatstat.explore)
library(viridis)

# Data-driven bandwidth
sig_default <- bw.diggle(X_ppp)
sig_default

sigmas <- c(sig_default, 500, 2000)
names(sigmas) <- c("bw.diggle", "500m", "2000m")

par(mfrow = c(1, 3), mar = c(2, 2, 3, 5))

for (i in seq_along(sigmas)) {

  kde_im <- density.ppp(
    X_ppp,
    sigma = sigmas[i],
    edge  = TRUE,
    at    = "pixels",
    dimyx = 256
  )

  plot(kde_im,
       main = paste("KDE (σ =", round(sigmas[i], 1), "m)"),
       col  = viridis(256))

  plot(X_ppp, add = TRUE, pch = 16, cex = 0.2)
}

par(mfrow = c(1, 1))


```
```{r}
library(spatstat.explore)
library(viridis)

# -----------------------------
# 1) Chosen bandwidth (automatic)
# -----------------------------
sig_default <- bw.diggle(X_ppp)
sig_default

kde_default <- density.ppp(
  X_ppp,
  sigma = sig_default,
  edge  = TRUE,
  at    = "pixels",
  dimyx = 256
)

plot(kde_default,
     main = paste("KDE of Firearm Incidents (σ =", round(sig_default, 1), "m)"),
     col  = viridis(256))
plot(X_ppp, add = TRUE, pch = 16, cex = 0.2)

# -----------------------------
# 2) KDE with σ = 500 m
# -----------------------------
kde_500 <- density.ppp(
  X_ppp,
  sigma = 500,
  edge  = TRUE,
  at    = "pixels",
  dimyx = 256
)

plot(kde_500,
     main = "KDE of Firearm Incidents (σ = 500 m)",
     col  = viridis(256))
plot(X_ppp, add = TRUE, pch = 16, cex = 0.2)

# -----------------------------
# 3) KDE with σ = 2000 m
# -----------------------------
kde_2000 <- density.ppp(
  X_ppp,
  sigma = 2000,
  edge  = TRUE,
  at    = "pixels",
  dimyx = 256
)

plot(kde_2000,
     main = "KDE of Firearm Incidents (σ = 2000 m)",
     col  = viridis(256))
plot(X_ppp, add = TRUE, pch = 16, cex = 0.2)

```


## RQ2: How do firearm violence rates vary across community districts, and are there statistically significant high-crime “hotspots” and low-crime “coldspots”?

```{r}
# =========================================================
# RQ2 Step 0: Outcome + Queen/Rook/kNN weights
# =========================================================
library(sf)
library(spdep)
library(dplyr)

# Outcome
y_var <- "rate_per_10k"
stopifnot(y_var %in% names(neigh_demo))

# Project for neighborhood structures
crs_m <- 26917
neigh_m <- neigh_demo |> st_transform(crs_m) |> st_make_valid()
geom <- st_geometry(neigh_m)
coords <- st_coordinates(st_centroid(geom))

# Outcome vector
y <- as.numeric(neigh_m[[y_var]])
stopifnot(length(y) == nrow(neigh_m))

# Build weights
nb_queen <- poly2nb(geom, queen = TRUE)
lw_queen <- nb2listw(nb_queen, style = "W", zero.policy = TRUE)

nb_rook  <- poly2nb(geom, queen = FALSE)
lw_rook  <- nb2listw(nb_rook,  style = "W", zero.policy = TRUE)

k <- 18
nb_knn_raw <- knn2nb(knearneigh(coords, k = k))
nb_knn <- make.sym.nb(nb_knn_raw)
lw_knn <- nb2listw(nb_knn, style = "W", zero.policy = TRUE)

# Diagnostics (like A3)
deg <- function(nb) sapply(nb, length)
cat("Avg degree — Queen:", mean(deg(nb_queen)),
    "| Rook:", mean(deg(nb_rook)),
    "| kNN:", mean(deg(nb_knn)), "\n")
cat("Connected components — Queen:", n.comp.nb(nb_queen)$nc,
    "| Rook:", n.comp.nb(nb_rook)$nc,
    "| kNN:", n.comp.nb(nb_knn)$nc, "\n")

# Compare global Moran I across structures (A3 Q3 style)
global_moran <- function(y, lw, label) {
  mt <- moran.test(y, lw, randomisation = FALSE, zero.policy = TRUE)
  tibble(
    weights = label,
    I  = unname(mt$estimate[["Moran I statistic"]]),
    EI = unname(mt$estimate[["Expectation"]]),
    p_value = mt$p.value
  )
}

global_tbl <- bind_rows(
  global_moran(y, lw_queen, "Queen"),
  global_moran(y, lw_rook,  "Rook"),
  global_moran(y, lw_knn,   paste0("kNN (k=", k, ")"))
)

print(global_tbl)

# =========================================================
# RQ2 Step 3: Plot neighborhood graphs (A3 style)
# =========================================================

plot(geom, border = "grey80", main = "Queen contiguity")
plot(nb_queen, coords, add = TRUE, col = "#d81b60")

plot(geom, border = "grey80", main = "Rook contiguity")
plot(nb_rook, coords, add = TRUE, col = "#1f78b4")

plot(geom, border = "grey80", main = paste0(k, "-NN"))
plot(nb_knn, coords, add = TRUE, col = "#33a02c")


```
```{r}
# Preferred weights for the rest of RQ2:
w <- lw_queen    # change to lw_queen or lw_knn if you prefer
w_nb <- nb_queen # matching nb object (needed for Gi* binary weights)

```

```{r}
# =========================================================
# RQ2 Q6a: Moran scatterplot (shooting rate per 10k)
# =========================================================
library(ggplot2)
library(ggrepel)

u <- y  # just to mirror A3 notation (u = outcome)
name_col <- if ("Neighbourh" %in% names(neigh_demo)) "Neighbourh" else "pretty_name"

# Local Moran (LISA)
lisa <- as.data.frame(localmoran(u, w, zero.policy = TRUE))

# robustly find p-value column
pcol <- grep("^Pr\\(", names(lisa), value = TRUE)[1]
names(lisa)[names(lisa) == pcol] <- "p_value"
if (!"Ii" %in% names(lisa)) names(lisa)[1] <- "Ii"

# Moran scatterplot pieces
z  <- as.numeric(scale(u))
Wz <- lag.listw(w, z, zero.policy = TRUE)
I_global <- unname(moran.test(u, w, randomisation = FALSE, zero.policy = TRUE)$estimate[["Moran I statistic"]])

df_scatter <- neigh_demo |>
  st_drop_geometry() |>
  transmute(
    name = .data[[name_col]],
    z = z,
    Wz = Wz
  ) |>
  bind_cols(lisa["p_value"]) |>
  mutate(
    quad = case_when(
      z >= 0 & Wz >= 0 ~ "HH",
      z <  0 & Wz <  0 ~ "LL",
      z >= 0 & Wz <  0 ~ "HL",
      TRUE             ~ "LH"
    ),
    sig = p_value < 0.05
  )

# Label significant HL/LH (spatial outliers); otherwise label top 6 residuals
lab_dat <- df_scatter |> filter(sig & quad %in% c("HL","LH"))
if (nrow(lab_dat) == 0) {
  lab_dat <- df_scatter |>
    mutate(resid = Wz - I_global * z) |>
    slice_max(order_by = abs(resid), n = 6)
}

ggplot(df_scatter, aes(x = z, y = Wz)) +
  geom_hline(yintercept = 0, linewidth = 0.6, linetype = 2, color = "grey40") +
  geom_vline(xintercept = 0, linewidth = 0.6, linetype = 2, color = "grey40") +
  geom_point(alpha = 0.8) +
  geom_abline(intercept = 0, slope = I_global, color = "firebrick", linewidth = 1) +
  geom_text_repel(
    data = lab_dat,
    aes(label = name),
    size = 3.2, min.segment.length = 0.1, box.padding = 0.3, max.overlaps = 30
  ) +
  labs(
    title = "Moran Scatterplot of Shooting Rate per 10k",
    subtitle = paste0("Global Moran’s I (slope) = ", sprintf("%.3f", I_global)),
    x = "Standardized shooting rate (z)",
    y = "Spatial lag Wz"
  ) +
  theme_minimal(base_size = 12)

```


```{r}
# =========================================================
# RQ2 Q6c: LISA quadrant map (sig vs non-sig)
# =========================================================
library(ggplot2)

# lisa_df MUST have the same ordering as neigh_demo / neigh_m used for weights
lisa_df <- tibble(
  demo_name = neigh_demo |> st_drop_geometry() |> pull(demo_name),
  Ii   = lisa_mat[, "Ii"],
  Z_Ii = lisa_mat[, "Z.Ii"],
  p_value = 2 * pnorm(-abs(lisa_mat[, "Z.Ii"]))
) |>
  mutate(p_adj = p.adjust(p_value, method = "holm"))

library(dplyr)
library(ggplot2)

lisa_sf <- neigh_demo |>
  mutate(z = z, Wz = Wz) |>
  left_join(lisa_df, by = "demo_name") |>
  mutate(
    sig = p_adj < 0.05,
    quad = case_when(
      z >= 0 & Wz >= 0 ~ "HH",
      z <  0 & Wz <  0 ~ "LL",
      z >= 0 & Wz <  0 ~ "HL",
      TRUE             ~ "LH"
    ),
    cluster = paste0(quad, ifelse(sig, " (sig)", " (ns)")),
    cluster = factor(cluster,
      levels = c("HH (sig)","LL (sig)","HL (sig)","LH (sig)",
                 "HH (ns)","LL (ns)","HL (ns)","LH (ns)")
    )
  )

pal <- c(
  "HH (sig)" = "#b2182b",
  "LL (sig)" = "#2166ac",
  "HL (sig)" = "#ef8a62",
  "LH (sig)" = "#67a9cf",
  "HH (ns)"  = "#cccccc",
  "LL (ns)"  = "#bfbfbf",
  "HL (ns)"  = "#d9d9d9",
  "LH (ns)"  = "#e6e6e6"
)

ggplot(lisa_sf) +
  geom_sf(aes(fill = cluster, alpha = ifelse(sig, 1, 0.35)),
          color = "white", linewidth = 0.2) +
  scale_fill_manual(values = pal, name = "LISA clusters") +
  scale_alpha_identity() +
  labs(
    title = "Local Moran’s I (LISA) for Shooting Rate per 10k",
    subtitle = "Holm-adjusted p < 0.05 shown in full opacity"
  ) +
  theme_void(base_size = 12) +
  theme(legend.position = "right")

```

```{r}
# =========================================================
# RQ2 Q7a: Getis-Ord Gi* scatterplot (binary weights)
# =========================================================
library(ggplot2)
library(ggrepel)

# Binary weights for Gi*
lw_bin <- nb2listw(w_nb, style = "B", zero.policy = TRUE)

GiZ <- as.numeric(localG(u, lw_bin, zero.policy = TRUE))

gstar_df <- tibble(
  name = neigh_demo |> st_drop_geometry() |> pull(.data[[name_col]]),
  z = z,
  GiZ = GiZ
) |>
  mutate(
    sig = abs(GiZ) > 1.96,
    cluster = case_when(
      GiZ >  1.96 ~ "Hot spot (High–High)",
      GiZ < -1.96 ~ "Cold spot (Low–Low)",
      TRUE ~ "Not significant"
    )
  )

ggplot(gstar_df, aes(x = z, y = GiZ, color = cluster)) +
  geom_hline(yintercept = c(-1.96, 1.96), linetype = "dashed", color = "grey40") +
  geom_point(size = 3, alpha = 0.9) +
  geom_text_repel(
    data = gstar_df |> filter(sig | abs(GiZ) > 2.5),
    aes(label = name),
    size = 3.2,
    max.overlaps = 60
  ) +
  scale_color_manual(values = c(
    "Hot spot (High–High)"  = "#d73027",
    "Cold spot (Low–Low)"  = "#4575b4",
    "Not significant"      = "grey70"
  )) +
  labs(
    x = "Standardized shooting rate (z)",
    y = "Local Getis–Ord Gi* (z-score)",
    color = "Cluster type",
    title = "Getis–Ord Gi* Hot- and Cold-Spot Detection (Shooting rate per 10k)"
  ) +
  theme_minimal(base_size = 14)

# Optional: list significant
gstar_df |> filter(sig) |> arrange(desc(GiZ))

```

```{r}
# =========================================================
# RQ2 Q7b: Gi* map (diverging palette; sig highlighted)
# =========================================================
library(ggplot2)

tor_gstar <- neigh_demo |>
  mutate(
    GiZ = GiZ,
    sig = abs(GiZ) > 1.96
  )

ggplot(tor_gstar) +
  geom_sf(aes(fill = GiZ, alpha = sig), color = "white", linewidth = 0.25) +
  scale_fill_distiller(type = "div", palette = "RdBu", direction = -1,
                       name = "Gi* (z-score)") +
  scale_alpha_manual(values = c(`TRUE` = 1, `FALSE` = 0.4), guide = "none") +
  labs(
    title = "Local Getis–Ord Gi* Hot and Cold Spots (Shooting rate per 10k)",
    subtitle = "Red = hot spots (high-rate clusters), Blue = cold spots (low-rate clusters)\nSignificant (|z|>1.96) shown in full opacity",
    caption = paste0("Weights: binary ", if (identical(w_nb, nb_rook)) "Rook" else if (identical(w_nb, nb_queen)) "Queen" else paste0("kNN (k=",k,")"))
  ) +
  theme_minimal(base_size = 13) +
  theme(panel.grid = element_blank(),
        legend.position = "right")

```

### RQ3: How are firearm violence rates associated with neighborhood socioeconomic characteristics after accounting for spatial dependence?

```{r}
# =========================================================
# RQ3 Setup: data + X's + Queen weights
# =========================================================
library(sf)
library(dplyr)
library(spdep)
library(spatialreg)
library(ggplot2)

# 1) Work in projected CRS for geometry operations (weights already built from polygons)
crs_m <- 26917
tor_sf <- neigh_demo |> st_transform(crs_m) |> st_make_valid()

# 2) Outcome
stopifnot("rate_per_10k" %in% names(tor_sf))
tor_sf$y <- tor_sf$rate_per_10k

# 3) Choose predictors (EDIT these names to match your columns)
#    Use proportions where appropriate.
#    Example: pct_visible_minority already a percent; can keep as % or /100, just be consistent.
stopifnot(all(c("median_total_income","unemploy_rate","pct_visible_minority","pct_black","population_total") %in% names(tor_sf)))

tor_sf <- tor_sf |>
  mutate(
    log_pop = log(population_total),
    vm_prop = pct_visible_minority / 100,
    black_prop = pct_black / 100
  )

# 4) Global formula used for all models
form_rq3 <- y ~ median_total_income + unemploy_rate + vm_prop + black_prop + log_pop

# 5) Spatial weights (Queen) from your RQ2 choice
lw <- lw_queen              # row-standardized for SAR models
nb <- nb_queen
lwB <- nb2listw(nb, style="B", zero.policy = TRUE)  # binary weights (often used for CAR + some diagnostics)

```
```{r}
map_resid <- function(sf_obj, resid_col, title){
  ggplot(sf_obj) +
    geom_sf(aes(fill = .data[[resid_col]]), color="white", linewidth=0.15) +
    scale_fill_viridis_c(option="magma") +
    labs(title = title, fill="Residual") +
    theme_minimal() +
    theme(panel.grid = element_blank())
}

```



```{r}
# =========================================================
# RQ3: Socioeconomic predictors + spatial models
# Outcome: shooting rate per 10k
# Models: OLS, SAR Error, SAR Lag, SAC, CAR
# Fix: impute missing covariates so ALL neighbourhoods are coloured
# =========================================================

suppressPackageStartupMessages({
  library(sf)
  library(dplyr)
  library(spdep)
  library(spatialreg)
  library(ggplot2)
  library(tibble)
})

# ---------------------------------------------------------
# 0) Setup: data, variables, CRS
# ---------------------------------------------------------
crs_m <- 26917
tor_sf <- neigh_demo |> st_transform(crs_m) |> st_make_valid()

stopifnot("rate_per_10k" %in% names(tor_sf))
tor_sf$y <- tor_sf$rate_per_10k

# REQUIRED predictors (edit if your names differ)
vars_raw <- c(
  "median_total_income",
  "unemploy_rate",
  "pct_visible_minority",
  "pct_black",
  "population_total"
)
stopifnot(all(vars_raw %in% names(tor_sf)))

# ---------------------------------------------------------
# 1) Impute missing predictors (MEDIAN) so we keep ALL rows
# ---------------------------------------------------------
tor_sf <- tor_sf |>
  mutate(across(
    all_of(vars_raw),
    ~ ifelse(is.na(.x), median(.x, na.rm = TRUE), .x)
  )) |>
  mutate(
    log_pop    = log(population_total),
    vm_prop    = pct_visible_minority / 100,
    black_prop = pct_black / 100
  )

# Global formula (same for every model)
form_rq3 <- y ~ median_total_income + unemploy_rate + vm_prop + black_prop + log_pop

# ---------------------------------------------------------
# 2) Spatial weights: Queen (build from SAME tor_sf)
# ---------------------------------------------------------
nb_queen  <- poly2nb(st_geometry(tor_sf), queen = TRUE)
lw_queen  <- nb2listw(nb_queen, style = "W", zero.policy = TRUE)  # SAR (row-standardized)
lwB_queen <- nb2listw(nb_queen, style = "B", zero.policy = TRUE)  # CAR (binary)

cat("RQ3 rows (kept):", nrow(tor_sf), "\n")

# ---------------------------------------------------------
# 3) Helpers: residual map + Moran test on residuals
# ---------------------------------------------------------
map_resid <- function(sf_obj, resid_col, title){
  ggplot(sf_obj) +
    geom_sf(aes(fill = .data[[resid_col]]), color = "white", linewidth = 0.15) +
    scale_fill_viridis_c(option = "magma") +
    labs(title = title, fill = "Residual") +
    theme_minimal(base_size = 12) +
    theme(panel.grid = element_blank())
}

# =========================================================
# (a) OLS
# =========================================================
mod_lm <- lm(form_rq3, data = st_drop_geometry(tor_sf))
print(summary(mod_lm))

tor_sf$lm_resid <- as.numeric(residuals(mod_lm))

moran_lm <- moran.test(tor_sf$lm_resid, lw_queen, zero.policy = TRUE)
print(moran_lm)

map_resid(tor_sf, "lm_resid", "OLS Residuals")

# =========================================================
# (b) SAR Spatial Error (SEM)
# =========================================================
mod_sem <- errorsarlm(
  formula = form_rq3,
  data    = st_drop_geometry(tor_sf),
  listw   = lw_queen,
  zero.policy = TRUE
)
print(summary(mod_sem))

tor_sf$sem_resid <- as.numeric(residuals(mod_sem))

moran_sem <- moran.test(tor_sf$sem_resid, lw_queen, zero.policy = TRUE)
print(moran_sem)

map_resid(tor_sf, "sem_resid", "SAR Error Residuals")

# =========================================================
# (c) SAR Spatial Lag (SLM)
# =========================================================
mod_slm <- lagsarlm(
  formula = form_rq3,
  data    = st_drop_geometry(tor_sf),
  listw   = lw_queen,
  zero.policy = TRUE
)
print(summary(mod_slm))

tor_sf$slm_resid <- as.numeric(residuals(mod_slm))

moran_slm <- moran.test(tor_sf$slm_resid, lw_queen, zero.policy = TRUE)
print(moran_slm)

map_resid(tor_sf, "slm_resid", "SAR Lag Residuals")

# =========================================================
# (d) SAR Spatial Lag–Error (SAC)
# =========================================================
mod_sac <- sacsarlm(
  formula = form_rq3,
  data    = st_drop_geometry(tor_sf),
  listw   = lw_queen,
  zero.policy = TRUE
)
print(summary(mod_sac))

tor_sf$sac_resid <- as.numeric(residuals(mod_sac))

moran_sac <- moran.test(tor_sf$sac_resid, lw_queen, zero.policy = TRUE)
print(moran_sac)

map_resid(tor_sf, "sac_resid", "SAR Lag–Error Residuals")

# =========================================================
# (e) CAR (spautolm, CAR family) — needs BINARY weights
# =========================================================
mod_car <- spautolm(
  formula = form_rq3,
  data    = st_drop_geometry(tor_sf),
  listw   = lwB_queen,
  family  = "CAR",
  zero.policy = TRUE,
  tol.solve = 1e-30
)
print(summary(mod_car))

tor_sf$car_resid <- as.numeric(residuals(mod_car))

# Moran’s I on CAR residuals: use the same binary weights
moran_car <- moran.test(tor_sf$car_resid, lwB_queen, zero.policy = TRUE)
print(moran_car)

map_resid(tor_sf, "car_resid", "CAR Residuals")

# =========================================================
# 4) Model comparison table (AIC)
# =========================================================
comp_tbl <- tibble(
  model = c("OLS", "SAR Error (SEM)", "SAR Lag (SLM)", "SAR Lag-Error (SAC)", "CAR"),
  AIC = c(AIC(mod_lm), AIC(mod_sem), AIC(mod_slm), AIC(mod_sac), AIC(mod_car))
) |>
  arrange(AIC)

print(comp_tbl)

# =========================================================
# 5) LM tests (HW4-style)
# Suggests whether lag or error is more appropriate
# =========================================================
lm_tests <- lm.LMtests(
  mod_lm,
  lw_queen,
  test = c("LMerr","LMlag","RLMerr","RLMlag"),
  zero.policy = TRUE
)
print(lm_tests)

```



